%% dmplex_geom.tex
%% Technical documentation of the distributed DMPlex geometry kernel in BioFEM-IA.
%% Self-contained: compile standalone or \input into a larger document.
%% Requires: amsmath, booktabs, listings, xcolor, hyperref, tikz.

\documentclass[11pt]{article}
\usepackage[margin=1.2in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,shapes}

\definecolor{codebg}{rgb}{0.97,0.97,0.97}
\definecolor{codekey}{rgb}{0.0,0.35,0.65}
\definecolor{codecomment}{rgb}{0.35,0.55,0.35}
\definecolor{codestring}{rgb}{0.65,0.10,0.10}

\lstset{
  language=C,
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codekey}\bfseries,
  commentstyle=\color{codecomment}\itshape,
  stringstyle=\color{codestring},
  breaklines=true,
  frame=single,
  framesep=4pt,
  numbers=left,
  numberstyle=\tiny\color{gray},
  numbersep=6pt,
  tabsize=2,
  showstringspaces=false,
  morekeywords={PetscErrorCode,PetscInt,PetscReal,PetscBool,PetscMPIInt,
                PetscSF,PetscSFNode,DM,IBMNodes,FE,
                DMPlexPatchLayout,DMPlexPatchExchange,DMPlexGeomCtx}
}

\title{\textbf{Distributed Geometry Kernel for Loop Subdivision Surfaces}\\
       \large Technical Reference: \texttt{dmplex\_geom.c} --- BioFEM-IA}
\author{Hossein Geshani}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
\section{Overview}
%% ============================================================

\texttt{dmplex\_geom.c} implements a parallel geometry evaluation pipeline
for triangulated immersed-boundary surfaces discretized with Loop subdivision
stencils.  The mesh is distributed across MPI ranks using PETSc's \textsc{DMPlex}
infrastructure.  Patch stencils that span partition boundaries are handled by
a \textsc{PetscSF} (Star Forest) coordinate exchange built once at initialization
and reused every geometry iteration.

The public API consists of four functions called in order:

\begin{center}
\begin{tabular}{lp{8cm}}
\toprule
Function & Purpose \\
\midrule
\texttt{FEM\_DMPlexGeomSetup(fem, comm)} & One-time setup: distribute mesh, build ownership table, layout, SF. \\
\texttt{FEM\_DMPlexGeomBuildNodeMap(fem)} & Build \texttt{ibm\_to\_global\_dof0} after Vecs are created. \\
\texttt{FEM\_DMPlexGeomUpdate(fem)} & Per-Newton-step: sync coordinates, recompute local geometry. \\
\texttt{FEM\_DMPlexGeomDestroy(fem)} & Free all context memory. \\
\bottomrule
\end{tabular}
\end{center}

All context is stored in \texttt{fem->geom\_ctx} (type \texttt{DMPlexGeomCtx}).

%% ============================================================
\section{Background Concepts}
%% ============================================================

\subsection{Loop Subdivision Stencils}

Loop subdivision refines a triangle mesh by inserting new vertices at edge midpoints
and repositioning existing vertices according to a weighted sum of their neighbours.
For a triangle $e$, the geometry at the subdivision limit surface requires coordinates
from a \emph{patch} of up to $k+6$ neighbouring vertices (where $k$ is the valence
of an irregular vertex) or 12 vertices for regular stencils.
These patch indices are precomputed and stored in \texttt{ibm->patch[16*e + i]},
using original vertex indices $\in [0, N_v)$.

\subsection{Replicated IBM Arrays vs Distributed PETSc Vecs}

A key architectural choice is that \texttt{IBMNodes} arrays (\texttt{x\_bp},
\texttt{nv1/nv2/nv3}, \texttt{patch}, \texttt{ire}, etc.) are \emph{fully replicated}
on every MPI rank.  For the LV mesh this costs only $608 \times (\text{few arrays})$
of memory per rank.  Because the mesh is generated analytically, every rank can
compute the same result independently --- no file I/O or broadcast is needed.

PETSc \emph{Vecs} (\texttt{Fint}, \texttt{xd}, etc.) are by contrast
\emph{distributed} (MPI Vecs): each rank owns a contiguous slice of the global DOF
range.  The two layers communicate via:
\begin{itemize}
  \item \textbf{Coordinate reads}: the exchange SF broadcasts updated coordinates from
        owning ranks into the local \texttt{ibm->x/y/z\_bp} arrays, keeping the
        replicated C arrays fresh.
  \item \textbf{Force writes}: \texttt{VecSetValues} with global DOF indices routes
        off-process contributions to their owning rank via PETSc's internal stash,
        flushed by \texttt{VecAssemblyBegin/End}.
\end{itemize}

\subsection{PETSc DMPlex: Chart, Height, Depth}

\textsc{DMPlex} represents mesh topology as a Directed Acyclic Graph (DAG) of
\emph{points}: cells, edges (if interpolated), and vertices.  All points on a rank
share a flat integer range called the \textbf{chart} $[\texttt{pStart}, \texttt{pEnd})$.

\textbf{Height} counts distance down from the top of the DAG; \textbf{depth} counts
up from the bottom:
\begin{center}
\begin{tabular}{lcc}
\toprule
Entity & Height & Depth \\
\midrule
Cell (triangle) & 0 & $d$ (mesh dim) \\
Edge & 1 & $d-1$ \quad (if interpolated) \\
Vertex & max & 0 \\
\bottomrule
\end{tabular}
\end{center}

BioFEM-IA uses an \emph{uninterpolated} mesh (no edges stored), so height~0 = cells,
height~1 = vertices, depth~0 = vertices.  After \texttt{DMPlexDistribute(overlap=0)},
each rank holds its own sub-DM with an independent chart.  The ranks have no shared
DM object; all cross-rank communication is explicit through PetscSF.

\textbf{Ghost vertices.}
With overlap=0, cells are partitioned with no ghost cell layers.  However,
a rank that owns triangle~$T$ must also hold all three of $T$'s vertices locally,
even if some of those vertices are shared with triangles on other ranks.  A
\emph{ghost vertex} is a vertex that is physically present in this rank's chart
(memory is allocated, coordinates are stored) but whose authoritative owner is
another rank.  Ghost vertices appear as \emph{leaves} in the DM's point SF
(\texttt{DMGetPointSF}): a leaf entry says ``my local point~$p$ is a copy of
rank~$r$'s point~$q$''.  Interior vertices (no sharing) are roots with no leaves.

\subsection{PetscSF: Star Forest}
\label{sec:sf}

A \texttt{PetscSF} encodes a point-to-many MPI communication pattern.

\begin{itemize}
  \item \textbf{Roots} --- authoritative data; each rank declares \texttt{nroots}
        root slots.  A root slot is simply an element of the root data array passed
        to \texttt{PetscSFBcast}.
  \item \textbf{Leaves} --- remote copies; each leaf is described by a
        \texttt{PetscSFNode\{rank, index\}} pointing to its root's location.
\end{itemize}

\texttt{PetscSFSetGraph(sf, nroots, nleaves, ilocal, ..., remote, ...)} defines the
graph: \texttt{remote[i].rank} and \texttt{remote[i].index} specify where leaf~$i$
fetches its data from.  The root data array has \texttt{nroots} elements; a leaf
pointing to \texttt{\{r, j\}} reads \texttt{rootdata[j]} on rank~$r$.

The two fundamental operations are:
\begin{align*}
\texttt{PetscSFBcast}:& \quad \text{root} \to \text{leaf}
  \quad (\text{owner sends, all needing ranks receive}) \\
\texttt{PetscSFReduce}:& \quad \text{leaf} \to \text{root}
  \quad (\text{contributions accumulated at owner})
\end{align*}

\textbf{Root index space.}
The exchange SF in this module sets \texttt{nroots = ibm->n\_v} on every rank,
even though each rank only owns $\approx N_v / P$ vertices.  This allows
\texttt{remote[i].index} to be the global IBM vertex index directly, so
\texttt{PetscSFBcast} reads \texttt{ibm->x\_bp[remote[i].index]} on the owning rank
without any remapping.  Unowned root slots are declared but never read.

Figure~\ref{fig:sf} illustrates the coordinate exchange SF.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  rank/.style={draw,rounded corners,minimum width=3.5cm,minimum height=1.8cm,fill=blue!8},
  root/.style={draw,circle,minimum size=0.55cm,fill=blue!20,font=\footnotesize},
  leaf/.style={draw,circle,minimum size=0.55cm,fill=orange!30,font=\footnotesize},
  arr/.style={-{Stealth},thick,blue!60},
  node distance=3.5cm
]
  \node[rank,label=above:\textbf{Rank 0}] (r0box) {};
  \node[rank,right=of r0box,label=above:\textbf{Rank 1}] (r1box) {};

  \node[root] (r0v0) at ([shift={(-1.0, 0.3)}]r0box.center) {$v_0$};
  \node[root] (r0v1) at ([shift={(0.0,  0.3)}]r0box.center) {$v_1$};
  \node[root] (r0v2) at ([shift={(1.0,  0.3)}]r0box.center) {$v_2$};
  \node[leaf] (r0lf) at ([shift={(0.0, -0.4)}]r0box.center) {$\ell_0$};

  \node[root] (r1v3) at ([shift={(-0.5, 0.3)}]r1box.center) {$v_3$};
  \node[root] (r1v4) at ([shift={(0.5,  0.3)}]r1box.center) {$v_4$};
  \node[leaf] (r1lf0) at ([shift={(-0.6,-0.4)}]r1box.center) {$\ell_0$};
  \node[leaf] (r1lf1) at ([shift={(0.6, -0.4)}]r1box.center) {$\ell_1$};

  \draw[arr] (r1v3) -- node[above,sloped,font=\scriptsize]{Bcast} (r0lf);
  \draw[arr] (r0v1) -- node[above,sloped,font=\scriptsize]{Bcast} (r1lf0);
  \draw[arr] (r1v4) -- node[above,sloped,font=\scriptsize]{Bcast} (r1lf1);

  \node[below=0.1cm of r0box,font=\footnotesize,text width=3.5cm,align=center]
    {owns $v_0,v_1,v_2$\\needs $v_3$ from R1};
  \node[below=0.1cm of r1box,font=\footnotesize,text width=3.5cm,align=center]
    {owns $v_3,v_4$\\needs $v_1$ from R0};
\end{tikzpicture}
\caption{Star Forest for coordinate exchange.  Blue circles are roots (owned vertices);
orange circles are leaves (remote coordinates needed by this rank's patch stencils).
Arrows show the Bcast direction.  Each leaf's \texttt{PetscSFNode} encodes the
owning rank and the global IBM vertex index.}
\label{fig:sf}
\end{figure}

%% ============================================================
\section{Data Structures}
%% ============================================================

\subsection{\texttt{DMPlexGeomCtx} --- the overall context}

All per-rank state is stored in \texttt{fem->geom\_ctx}.

\begin{center}
\begin{tabular}{lll}
\toprule
Field & Size / Type & Description \\
\midrule
\texttt{comm}              & \texttt{MPI\_Comm}    & Communicator (usually \texttt{PETSC\_COMM\_WORLD}). \\
\texttt{initialized}       & \texttt{PetscBool}    & Set true after \texttt{FEM\_DMPlexGeomSetup}. \\
\texttt{layout}            & \texttt{DMPlexPatchLayout} & Local cell stencil data (see \S\ref{sec:layout}). \\
\texttt{exchange}          & \texttt{DMPlexPatchExchange} & Coordinate SF and buffers (see \S\ref{sec:exchange_struct}). \\
\texttt{origVert}          & $N_\text{local verts}$ & DM local vertex $\to$ global IBM vertex index. \\
\texttt{ownerRank}         & $N_v$ (replicated)    & MPI rank owning each global IBM vertex. \\
\texttt{nLocalVerts}       & scalar                & \texttt{vEnd - vStart} in this rank's DM. \\
\texttt{nOwnedVerts}       & scalar                & Vertices owned by this rank; Vec local size. \\
\texttt{ibm\_to\_local\_idx}  & $N_v$ (mostly $-1$) & Global IBM vertex $\to$ local Vec index ($-1$ if unowned). \\
\texttt{ibm\_to\_global\_dof0}& $N_v$ (replicated)  & Global IBM vertex $\to$ first global DOF in MPI Vec. \\
\bottomrule
\end{tabular}
\end{center}

\textbf{\texttt{ibm\_to\_local\_idx}.}
This array has $N_v$ entries on every rank but most are $-1$.  The non-negative
entries form a compact sequential mapping $[0, \texttt{nOwnedVerts})$ over the
vertices this rank owns.  \texttt{VecCreateMPI} is called with
\texttt{dof * nOwnedVerts} as the local size, so local Vec index $i$ corresponds
to global IBM vertex \texttt{v} where \texttt{ibm\_to\_local\_idx[v] = i}.

\textbf{\texttt{ibm\_to\_global\_dof0}.}
Built by \texttt{FEM\_DMPlexGeomBuildNodeMap} \emph{after} Vecs exist.  Entry
\texttt{ibm\_to\_global\_dof0[v]} is the global DOF index of the $x$-component
of vertex~$v$ in any MPI Vec.  Used by \texttt{FInternalAct} to call
\texttt{VecSetValues} with correct global indices when assembling \texttt{Fint}.

\subsection{\texttt{DMPlexPatchLayout}}
\label{sec:layout}

Per-rank local copy of subdivision stencil data.  All arrays indexed by
local cell index $\ell_c \in [0, N_\text{local})$.

\begin{center}
\begin{tabular}{lll}
\toprule
Field & Size & Description \\
\midrule
\texttt{dm}        & ---                        & Ref-counted distributed DM (this rank's sub-mesh). \\
\texttt{orig\_cell}& $N_\text{local}$           & Original IBM cell index $e$ for $\ell_c$; $-1$ for unknown. \\
\texttt{ire}       & $N_\text{local}$           & 0 = regular stencil, 1 = irregular. \\
\texttt{irv}       & $N_\text{local}$           & Irregular-vertex flag. \\
\texttt{val}       & $N_\text{local}$           & Valence of the irregular vertex. \\
\texttt{patch}     & $16 \times N_\text{local}$ & Original vertex indices of the stencil. Unused: \texttt{PATCH\_MISSING}. \\
\texttt{nLocalCells}  & ---                     & $N_\text{local} = \texttt{cEnd} - \texttt{cStart}$. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{\texttt{DMPlexPatchExchange}}
\label{sec:exchange_struct}

Holds the SF and coordinate receive buffers.  All arrays indexed by remote node $i
\in [0, N_\text{remote})$.

\begin{center}
\begin{tabular}{lll}
\toprule
Field & Size & Description \\
\midrule
\texttt{sf}           & ---               & PetscSF: \texttt{nroots=$N_v$}, \texttt{nleaves=$N_\text{remote}$}. \\
\texttt{nRemoteNodes} & ---               & Unique cross-rank patch vertices needed by this rank. \\
\texttt{orig\_node}   & $N_\text{remote}$ & Global IBM vertex index of remote node $i$. \\
\texttt{owner\_rank}  & $N_\text{remote}$ & MPI rank owning \texttt{orig\_node[$i$]}. \\
\texttt{owner\_index} & $N_\text{remote}$ & Root slot on owner (= \texttt{orig\_node[$i$]}). \\
\texttt{x\_remote}    & $N_\text{remote}$ & Receive buffer for $x$ coordinates after Bcast. \\
\texttt{y\_remote}    & $N_\text{remote}$ & Receive buffer for $y$ coordinates. \\
\texttt{z\_remote}    & $N_\text{remote}$ & Receive buffer for $z$ coordinates. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Memory layout summary}

Table~\ref{tab:memory} summarises which arrays are replicated across all ranks
and which are per-rank.

\begin{table}[h]
\centering
\begin{tabular}{llll}
\toprule
Array & Size on each rank & Nature & Lives in \\
\midrule
\texttt{ibm->x\_bp[]} & $N_v = 608$ (full) & Replicated C array, kept fresh by SF & \texttt{IBMNodes} \\
\texttt{ibm->nv1/nv2/nv3[]} & $N_e = 1152$ (full) & Replicated & \texttt{IBMNodes} \\
\texttt{ibm->patch[16*$N_e$]} & full & Replicated & \texttt{IBMNodes} \\
\texttt{act->elem\_act\_data[]} & $N_e$ (full) & Replicated & \texttt{ActData} \\
\texttt{ctx->ownerRank[$N_v$]} & full & Replicated (built by Allreduce) & \texttt{DMPlexGeomCtx} \\
\texttt{ctx->ibm\_to\_local\_idx[$N_v$]} & full, mostly $-1$ & Replicated & \texttt{DMPlexGeomCtx} \\
\texttt{ctx->ibm\_to\_global\_dof0[$N_v$]} & full & Replicated (built by Allreduce) & \texttt{DMPlexGeomCtx} \\
\midrule
\texttt{ctx->origVert[]} & $\approx N_v/P + \text{bdry}$ & Per-rank DM vertex map & \texttt{DMPlexGeomCtx} \\
\texttt{layout.orig\_cell[]} & $\approx N_e/P$ & Per-rank cell map & \texttt{DMPlexPatchLayout} \\
\texttt{layout.patch[16*$N_\text{local}$]} & $\approx 16 N_e/P$ & Per-rank stencil slice & \texttt{DMPlexPatchLayout} \\
\texttt{exchange.orig\_node[]} & $N_\text{remote}$ & Per-rank remote fetch list & \texttt{DMPlexPatchExchange} \\
\texttt{exchange.x/y/z\_remote[]} & $N_\text{remote}$ & Per-rank receive buffers & \texttt{DMPlexPatchExchange} \\
\midrule
\texttt{fem->Fint} (MPI Vec) & $\texttt{dof} \times N_\text{owned}$ & Distributed & PETSc MPI Vec \\
\texttt{fem->xd} (MPI Vec) & $\texttt{dof} \times N_\text{owned}$ & Distributed & PETSc MPI Vec \\
\bottomrule
\end{tabular}
\caption{Memory layout across MPI ranks for the LV mesh ($N_v=608$, $N_e=1152$, $P=8$).
Per-rank sizes are approximate.}
\label{tab:memory}
\end{table}

%% ============================================================
\section{Setup Pipeline: \texttt{FEM\_DMPlexGeomSetup}}
%% ============================================================

The one-time setup is orchestrated by \texttt{FEM\_DMPlexGeomSetup(fem, comm)}.
It calls four internal functions in sequence, then builds \texttt{ibm\_to\_local\_idx}
directly from the resulting ownership table.  Figure~\ref{fig:pipeline} summarises
the pipeline.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  box/.style={draw,rounded corners,fill=blue!8,minimum width=5.8cm,
              minimum height=0.75cm,font=\small,align=center},
  arr/.style={-{Stealth},thick},
  note/.style={font=\footnotesize,text=gray,text width=5.2cm,align=left}
]
\node[box] (s1) {\texttt{CreateDistributedDMPlex\_()}};
\node[box,below=0.5cm of s1] (s2) {\texttt{BuildOriginalVertexOwnerTable\_()}};
\node[box,below=0.5cm of s2] (s3) {\texttt{DMPlexPatchLayoutCreate\_()}};
\node[box,below=0.5cm of s3] (s4) {\texttt{BuildPatchCoordinateExchange\_()}};
\node[box,below=0.5cm of s4] (s5) {Build \texttt{ibm\_to\_local\_idx}, set \texttt{nOwnedVerts}};

\draw[arr] (s1)--(s2);
\draw[arr] (s2)--(s3);
\draw[arr] (s3)--(s4);
\draw[arr] (s4)--(s5);

\node[note,right=0.9cm of s1] {Build serial DM on R0 (no edges), distribute overlap=0; extract \texttt{origCell[]}/\texttt{origVert[]} from migration SF.};
\node[note,right=0.9cm of s2] {Ghost detection via point SF; \texttt{MPI\_Allreduce(MAX)} $\Rightarrow$ \texttt{ownerRank[0..N\_v)}.};
\node[note,right=0.9cm of s3] {Copy \texttt{ire/irv/val/patch} from global \texttt{ibm} to local layout. No communication.};
\node[note,right=0.9cm of s4] {Identify cross-rank patch nodes; build PetscSF. $\mathcal{O}(N/P)$.};
\node[note,right=0.9cm of s5] {Walk \texttt{origVert}: skip ghosts, assign sequential Vec index to owned verts.};
\end{tikzpicture}
\caption{One-time setup pipeline inside \texttt{FEM\_DMPlexGeomSetup}.}
\label{fig:pipeline}
\end{figure}

\subsection{\texttt{BuildSerialDMPlex\_}}

Rank~0 fills the cell connectivity and coordinate arrays from \texttt{ibm} and
passes them to \texttt{DMPlexCreateFromCellListPetsc} with
\texttt{interpolate=PETSC\_FALSE}.  All other ranks pass empty arrays.
Skipping edge interpolation avoids PETSc's $\mathcal{O}(N^{3+})$ serial
edge-building path; distribution of an uninterpolated mesh at \texttt{overlap=0}
is fully supported.

\subsection{\texttt{CreateDistributedDMPlex\_}}
\label{sec:createdist}

Calls \texttt{DMPlexDistribute(dm, 0, \&migrationSF, \&distributed)}.
The migration SF records, for each local point on each rank, the serial point on
rank~0 it originated from.  Both \texttt{origCell[]} and \texttt{origVert[]} are
extracted from this SF in a single $\mathcal{O}(n_\text{leaves})$ pass:
\begin{itemize}
  \item For each leaf $i$: local point \texttt{ilocal[i]}, serial origin \texttt{iremote[i].index}.
  \item If the local point is a cell: \texttt{origCell[lc] = iremote[i].index - serial\_cStart}.
  \item If it is a vertex: \texttt{origVert[lv] = iremote[i].index - serial\_vStart}.
\end{itemize}
For \texttt{np=1} the distributed DM is \texttt{NULL}; the identity mapping is applied directly.

\subsection{\texttt{BuildOriginalVertexOwnerTable\_}}
\label{sec:owner}

Produces \texttt{ownerRank[i]} = MPI rank that owns original vertex~$i$,
consistent across all ranks.

\textbf{Step 1: Ghost detection.}
The DM's point SF (\texttt{DMGetPointSF}) has ghost points as its leaves.
The leaf's \texttt{ilocal[i]} gives the local chart index of the ghost copy:
\[
  \texttt{isGhost}[p - p_\text{start}] = \texttt{true}
  \quad \forall p \in \{\text{SF leaf local indices}\}
\]

\textbf{Step 2: Local ownership claim.}
Walk all local DM vertices $v \in [v_\text{start}, v_\text{end})$.  Read
\texttt{orig = origVert[v - vStart]}.  For non-ghost vertices only, set
\texttt{localOwner[orig] = myRank}.  Ghost vertices (shared partition-boundary
copies owned by another rank) are skipped.

\textbf{Step 3: Global reduction.}
\[
  \texttt{ownerRank[i]} = \max_{\text{all ranks}} \texttt{localOwner[i]}
\]
Non-owners hold $-1$; the one owning rank holds its rank ID~$\geq 0$.
\texttt{MPI\_MAX} selects the unique owner.

\subsection{\texttt{DMPlexPatchLayoutCreate\_}}

Iterates over locally owned cells $lc \in [0, N_\text{local})$, reads
$e = \texttt{origCell}[lc]$, and copies \texttt{ibm->ire[e]}, \texttt{irv[e]},
\texttt{val[e]}, and \texttt{patch[16e+i]} into the layout arrays.
No communication; each rank reads from its local replicated \texttt{ibm} copy.

\subsection{\texttt{BuildPatchCoordinateExchange\_}}
\label{sec:exchange}

Determines which original vertices this rank needs from others and builds
the PetscSF.

\subsubsection{Building \texttt{isLocalOrig}}

Repeats the ghost-detection walk from \S\ref{sec:owner} to build
\texttt{isLocalOrig[i] = true} iff original vertex~$i$ is a non-ghost vertex on
this rank (i.e.\ this rank owns it and its \texttt{ibm->x\_bp[i]} is authoritative).

\subsubsection{Main loop --- collect remote nodes}

\begin{lstlisting}
alreadyAdded[n_v] = {false}    /* dedup: each vertex added at most once */

for lc = 0..nLocalCells:
    nen = 12 if regular, else val+6
    for i = 0..nen:
        node = layout->patch[16*lc + i]
        if MISSING or >= n_v: continue   /* empty / auxiliary slot */
        if isLocalOrig[node]: continue   /* I own it — always fresh */
        if alreadyAdded[node]: continue  /* already queued */

        /* node is needed but owned by another rank */
        orig_node[nRemoteNodes]  = node
        owner_rank[nRemoteNodes] = ownerRank[node]
        nRemoteNodes++
        alreadyAdded[node] = true
\end{lstlisting}

The two $\mathcal{O}(1)$ hash-array lookups (\texttt{isLocalOrig},
\texttt{alreadyAdded}) replace the former $\mathcal{O}(N_v/P)$ linear scans,
reducing the overall complexity from $\mathcal{O}(N^2/P^2)$ to $\mathcal{O}(N/P)$.

\subsubsection{SF construction}

\begin{lstlisting}
for i = 0..nRemoteNodes:
    remote[i].rank  = owner_rank[i]
    remote[i].index = orig_node[i]   /* global IBM vertex index */

PetscSFSetGraph(sf,
    ibm->n_v,      /* nroots:  full N_v index space — no remapping needed */
    nRemoteNodes,  /* nleaves: unique cross-rank vertices on this rank */
    NULL,          /* ilocal:  leaves at x_remote[0,1,...] sequentially */
    remote);       /* PetscSFNode: {ownerRank, origVertIndex} per leaf */
\end{lstlisting}

Root slots that no leaf ever points to are never touched during \texttt{PetscSFBcast}.

\subsubsection{Complexity comparison}

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
& Naive & Optimised \\
\midrule
Per patch-node locality check & $\mathcal{O}(N_v/P)$ linear scan & $\mathcal{O}(1)$ array lookup \\
Per patch-node dedup check    & $\mathcal{O}(N_\text{remote})$ scan & $\mathcal{O}(1)$ flag \\
Full loop & $\mathcal{O}(N^2/P^2)$ & $\mathcal{O}(N/P)$ \\
Extra memory & 0 & $2 \times N_v$ booleans \\
\bottomrule
\end{tabular}
\caption{Complexity of \texttt{BuildPatchCoordinateExchange\_}.}
\label{tab:complexity}
\end{table}

\subsection{Building \texttt{ibm\_to\_local\_idx} and \texttt{nOwnedVerts}}

Immediately after the four sub-functions, \texttt{FEM\_DMPlexGeomSetup} builds
the Vec local-index map:

\begin{lstlisting}
ctx->nOwnedVerts = 0;
for lv = 0..nLocalVerts:
    v = origVert[lv]
    if v >= 0 && v < n_v && ownerRank[v] == myrank:
        ibm_to_local_idx[v] = nOwnedVerts++
\end{lstlisting}

Ghost vertices in \texttt{origVert} (those with \texttt{ownerRank[v] != myrank})
are skipped, so each global vertex gets exactly one owner's sequential index.
The sum $\sum_P \texttt{nOwnedVerts} = N_v$ exactly, which is the invariant
required by \texttt{VecCreateMPI}.

%% ============================================================
\section{Node Map: \texttt{FEM\_DMPlexGeomBuildNodeMap}}
%% ============================================================

Called \emph{after} \texttt{InitVecs} (which creates the MPI Vecs), this function
builds \texttt{ibm\_to\_global\_dof0} --- the bridge from global IBM vertex indices
to global PETSc Vec DOF indices.

\begin{lstlisting}
VecGetOwnershipRange(fem->Fint, &lo, &hi);  /* this rank's DOF range [lo, hi) */
oi = 0;
for lv = 0..nLocalVerts:
    v = origVert[lv]
    if ownerRank[v] != myrank: continue
    ibm_to_global_dof0[v] = lo + oi * dof
    oi++

MPI_Allreduce(MPI_IN_PLACE, ibm_to_global_dof0, n_v, MPI_INT, MPI_MAX, comm)
\end{lstlisting}

Each rank fills only the entries for its owned vertices (using its local DOF offset
\texttt{lo}).  \texttt{MPI\_Allreduce(MAX)} distributes the complete table to every
rank.  Result: \texttt{ibm\_to\_global\_dof0[v]} holds the global index of the
$x$-DOF for vertex~$v$ in any MPI Vec, on every rank.

%% ============================================================
\section{Per-Iteration Kernel}
%% ============================================================

Every Newton step calls \texttt{FEM\_DMPlexGeomUpdate(fem)}, which:

\begin{enumerate}
  \item \textbf{Syncs coordinates.}
        \texttt{UpdatePatchRemoteCoordinates\_} fires three sequential SF broadcasts:
        \begin{align*}
          \texttt{x\_remote}[i] &\leftarrow \texttt{ibm->x\_bp[orig\_node[i]]}
          \text{ on rank \texttt{owner\_rank[i]}} \\
          \texttt{ibm->x\_bp[orig\_node[i]]} &\leftarrow \texttt{x\_remote}[i]
        \end{align*}
        (repeated for $y$, $z$).  After this call, every rank's \texttt{ibm->x/y/z\_bp}
        is fresh for all patch-stencil vertices.

  \item \textbf{Recomputes local geometry.}
        Loops over \texttt{lc = 0..nLocalCells}, calls \texttt{ElemUpdateGeomSubdiv(fem, ec)}
        for each owned cell.
\end{enumerate}

%% ============================================================
\section{Parallel Force Assembly: \texttt{FInternalAct}}
%% ============================================================

\texttt{FInternalAct} assembles the internal force vector \texttt{Fint} in parallel
using \texttt{VecSetValues}.

\begin{lstlisting}
for lc = 0..nLocalCells:
    ec = layout.orig_cell[lc]          /* global IBM cell index */
    ElemUpdFint(fem, ec, Fb)           /* compute elem force, size nen*dof */

    for i = 0..nen:
        node  = ibm->patch[16*ec + i]  /* global IBM vertex index */
        gdof0 = ibm_to_global_dof0[node]   /* first global DOF */
        dofs = {gdof0, gdof0+1, gdof0+2}
        vals = {Fb[3i], Fb[3i+1], Fb[3i+2]}
        VecSetValues(Fint, 3, dofs, vals, ADD_VALUES)

VecAssemblyBegin(Fint)
VecAssemblyEnd(Fint)
\end{lstlisting}

\textbf{\texttt{VecSetValues} with global indices.}
\texttt{Fint} is an MPI Vec; each rank owns global DOFs $[\texttt{lo}, \texttt{hi})$.
When \texttt{VecSetValues} is called with a global index:
\begin{itemize}
  \item In $[\texttt{lo}, \texttt{hi})$: written directly to the local buffer.
  \item Outside $[\texttt{lo}, \texttt{hi})$: buffered in PETSc's internal
        \emph{stash} (no MPI yet).
\end{itemize}

A patch stencil crosses partition boundaries: vertex~$v$ may be owned by rank~1
while the cell is on rank~0.  Rank~0 computes the force contribution and calls
\texttt{VecSetValues} at $v$'s global DOF --- that entry enters rank~0's stash.

\textbf{\texttt{VecAssemblyBegin/End}.}
\texttt{VecAssemblyBegin} initiates MPI sends of all stashed off-process entries.
\texttt{VecAssemblyEnd} waits for receives and applies \texttt{ADD\_VALUES} to the
owned local buffer.  After \texttt{VecAssemblyEnd}, every rank's owned slice of
\texttt{Fint} contains the correct total force accumulated from contributions
by all ranks whose stencils touched those DOFs.

There is no fixed limit on the number of off-process entries per
\texttt{VecSetValues} call; PETSc's stash grows dynamically.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  box/.style={draw,rounded corners,fill=blue!8,minimum width=5cm,
              minimum height=0.7cm,font=\small,align=center},
  arr/.style={-{Stealth},thick},
  note/.style={font=\footnotesize,text=gray,text width=5cm,align=left}
]
\node[box] (a) {\texttt{FEM\_DMPlexGeomUpdate}};
\node[box,below=0.4cm of a] (b) {\texttt{FInternalPreCalc}};
\node[box,below=0.4cm of b] (c) {loop: \texttt{ElemUpdFint} + \texttt{VecSetValues}};
\node[box,below=0.4cm of c] (d) {\texttt{VecAssemblyBegin/End}};

\draw[arr](a)--(b); \draw[arr](b)--(c); \draw[arr](c)--(d);

\node[note,right=0.8cm of a] {Coord sync: SF Bcast $\times 3$ (x,y,z).};
\node[note,right=0.8cm of b] {Pre-compute kinematics for local cells.};
\node[note,right=0.8cm of c] {Local compute; off-proc DOFs go to stash.};
\node[note,right=0.8cm of d] {Flush stash via MPI; ADD\_VALUES at owners.};
\end{tikzpicture}
\caption{Force assembly pipeline per Newton step.}
\label{fig:assembly}
\end{figure}

%% ============================================================
\section{Timing and Benchmarks}
%% ============================================================

\texttt{CreateDistributedDMPlex\_} records wall-clock times via \texttt{MPI\_Wtime}.
Selected results from the mesh sweep (GRACE cluster, \texttt{overlap=0},
no serial interpolation):

\begin{center}
\begin{tabular}{rrccc}
\toprule
$N_x \times N_y$ & $N_\text{elmt}$ & $P$ & Setup (s) & Iter-1 (s) \\
\midrule
$16\times12$   &    334        & 1  & 0.0043  & 0.00650 \\
$64\times48$   &  5\,926       & 4  & 0.012   & 0.02808 \\
$64\times48$   &  5\,926       & 32 & 0.030   & 0.00359 \\
$128\times96$  & 24\,132       & 8  & 0.029   & 0.06522 \\
$128\times96$  & 24\,132       & 32 & 0.015   & 0.01513 \\
$256\times192$ & 97\,414       & 32 & 0.058   & 0.05829 \\
$512\times384$ & 391\,434      & 32 & 0.24    & 0.25191 \\
$640\times480$ & 612\,166      & 32 & 0.38    & 0.41079 \\
\bottomrule
\end{tabular}
\end{center}

Setup is now dominated by \texttt{DMPlexDistribute} rather than
\texttt{BuildPatchCoordinateExchange\_}, which was the former $\mathcal{O}(N^2/P^2)$
bottleneck.  Iteration time scales near-linearly with $N/P$.

%% ============================================================
\section{Call Graph Summary}
%% ============================================================

\begin{lstlisting}[language={},numbers=none,frame=none,backgroundcolor=\color{white}]
main()
  FEM_DMPlexGeomSetup(fem, PETSC_COMM_WORLD)          /* one-time setup */
    CreateDistributedDMPlex_()
      BuildSerialDMPlex_()        /* rank 0 only, PETSC_FALSE = no edges */
      DMPlexDistribute(overlap=0) /* collective; migration SF -> origCell[], origVert[] */
    BuildOriginalVertexOwnerTable_(origVert)
      DMGetPointSF()              /* ghost detection */
      MPI_Allreduce(MPI_MAX)      /* collective, size N_v -> ownerRank[] */
    DMPlexPatchLayoutCreate_(origCell)  /* local, no comms */
    BuildPatchCoordinateExchange_()
      PetscSFSetGraph + PetscSFSetUp   /* build exchange SF */
    [inline] build ibm_to_local_idx, nOwnedVerts

  InitVecs(fem)                         /* VecCreateMPI(dof*nOwnedVerts, dof*N_v) */

  FEM_DMPlexGeomBuildNodeMap(fem)       /* after Vecs exist */
    VecGetOwnershipRange()
    MPI_Allreduce(MPI_MAX)             /* -> ibm_to_global_dof0[] on all ranks */

  [per Newton step:]
  FEM_DMPlexGeomUpdate(fem)
    UpdatePatchRemoteCoordinates_()
      PetscSFBcast() x3               /* collective: x, y, z coords */
    ElemUpdateGeomSubdiv() per lc     /* local */

  FInternalAct(fem)
    ElemUpdFint() per lc              /* local */
    VecSetValues(ADD_VALUES)          /* local + stash for off-proc */
    VecAssemblyBegin/End()            /* collective: flush stash */

  FEM_DMPlexGeomDestroy(fem)          /* free ctx */
\end{lstlisting}

%% ============================================================
\section{Runtime Options}
%% ============================================================

\begin{center}
\begin{tabular}{lll}
\toprule
Option & Default & Description \\
\midrule
\texttt{-nbody} & 1 & Number of IBM bodies to process. \\
\bottomrule
\end{tabular}
\end{center}

The ghost overlap is hardcoded to \texttt{overlap=0}.  Cross-boundary stencil nodes
are handled entirely by the PetscSF coordinate exchange; no ghost cell topology
is needed or created.

\end{document}
